[Re] On the coexistence of specialists and generalists

### 2018-05-08 article/Poisot-2017.tex
Update gitignore
Compile the paper

\documentclass[10pt,a4paper,onecolumn]{article}
% \usepackage[utf8]{inputenc}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{hyperref}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor=
{
      Timothée Poisot,
  },
            pdftitle=
{
[Re] On the coexistence of specialists and generalists
},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}
\usepackage{tcolorbox}
\usepackage{ragged2e}
\usepackage{fontspec}
\usepackage{fontawesome}
\usepackage{amssymb,amsmath}
\usepackage{caption}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

@@

\section{Introduction}\label{introduction}

The coexistence of specialists and generalist within ecological
communities is a long-standing question. \textcite{wils94csg} have
suggested that this coexistence can be understood when examined in the
light of (i) differential fitness loss associated to specialism, (ii)
active habitat selection, (iii) negative density dependence due to
competition, and (iv) stochastic changes in habitat quality, that allow
combinations of species to persist even though coexistence would not be
possible in a purely deterministic world. Here I propose an
implementation of this model in \emph{Julia} \autocite{beza17jfa}, and
show that it is able to reproduce most figures from the original
manuscript.

\section{Methods}\label{methods}

@@

In this matrix, \(K_1\) is the quality of habitat 1, \(K_2\) is the
quality of habitat 2, \(a\) is the fitness cost of the specialist in its
non-optimal environment, and \(b\) is the fitness cost of generalism.
Note that \(1 > b > a > 0\).

@@

The value of \(p_l\) for which \(D_{l1}=D_{l2}\) is

@@

We fix \(p_m\) and \(p_n\), and find the value of \(p_l\), while
enforcing the constraint of \(0 \leq p_l \leq 1\). Repeating this
procedure a few times for the different species yields the optimal
values of \(\mathbf{p}\); we can measure the density of individuals in
both habitats. Before we do so, there is a proportion \(g\) of
individuals that select habitats at random. Given a total population
size of \(N_i\), there are \(N_i(g/2)\) individuals will go to either
habitat, and \(N_i(1-g)p_i\) will pick habitat 1. With this information,
we can write the matrix describing habitat selection:

@@

@@

\section{Results}\label{results}

Original sources were not available, and no attempts were made to
contact the authors. For some non-stochastic situations, it is possible
to calculate expected values by hand. The original manuscript does
provide some of these values, and they were used to test this
implementation. All default parameter values are given in
~\ref{tbl:prm}. In all figures, circles represent the generalist, upper
and lower triangles represent the specialists of habitats 1 and 2
(respectively), and a diamond is used when both specialists have the
same behaviour.

Figure 2A and 2B in the original manuscript provide a good diagnostic
value, as they are based on non-stochastic situations. These are
reproduced in ~\ref{fig:02}. As in the original simulations, the
generalist alone overshoots its carrying capacity at \(t=5\), and the
stabilizes at a value of \(N_3^\star = 270\). Note also that
\(p_3 = 2/3\) , which represents the ratio of habitat qualities. When
adding specialists at initial densities of \(N_1 = N_2 = 1\), we observe
(i) that the preference of the generalist shifts towards habitat 1 to
avoid competition in the smallest habitat, and (ii) the density of the
specialist of habitat 2 stabilizes as soon as the generalist abandons
habitat 2 (at \(t \approx 35\)).

\begin{figure}
\centering
\includegraphics{figure02.pdf}
\caption{Population dynamics and habitat preference of the generalist
alone (left), and following invasion by the two specialists at the
generalist equilibrium (right).}\label{fig:02}
\end{figure}

In ~\ref{fig:03}, this implementation gives the same qualitative results
as the original article. There are small quantitative differences in
output, which are in my opinion explained by the initial population
densities (which are not given in the original article). Absent
variation in habitat quality, generalists cannot persist, and the
specialists both reach \(N^\star = \bar K\). There is a value of
variation for which the densities of specialists and generalists are
equal (about 110 for tied variations, and 130 for independent
variations).

\begin{figure}
\centering
\includegraphics{figure03.pdf}
\caption{Consequences of increasing the range of variation of both
\(K_1\) and \(K_2\). Dashed lines represent independent variation, and
solid lines tied variations, as in the original article.}\label{fig:03}
\end{figure}

In ~\ref{fig:04}, the same qualitative result (generalists increase in
abundance when one habitat becomes larger). The quantitative differences
can, again, be explained by changes in the initial population densities.
Finally, the results in ~\ref{fig:05} and ~\ref{fig:06} can also be
replicated.

\begin{figure}
\centering
\includegraphics{figure04.pdf}
\caption{Consequences of increasing the range of variation of both
\(K_1\) and \(K_2\). Dashed lines represent independent variation, and
solid lines tied variations, as in the original article.}\label{fig:04}
\end{figure}

\begin{figure}
\centering
\includegraphics{figure05.pdf}
\caption{Consequences of increasing the range of variation of both
\(K_1\) and \(K_2\). Dashed lines represent independent variation, and
solid lines tied variations, as in the original article.}\label{fig:05}
\end{figure}

\begin{figure}
\centering
\includegraphics{figure06.pdf}
\caption{Consequences of increasing the range of variation of both
\(K_1\) and \(K_2\). Dashed lines represent independent variation, and
solid lines tied variations, as in the original article.}\label{fig:06}
\end{figure}

\section{Conclusion}\label{conclusion}

I believe that this implementation faithfully reproduces the results of
\textcite{wils94csg}. A single run of the model (using 100 generations
as in the original article) completes in
\(\approx 5\times 10^{-3} \text{s}\). This makes this implementation
usable for teaching, as it is not time consuming to generate results. As
an example, I merged the output of ~\ref{fig:03} and ~\ref{fig:05}, to
generate the response of Shannon's evenness (which measures the extent
to which all species have the same density) to changes in the rate of
variation \emph{and} generalist ability. The results are presented in
~\ref{fig:surf}.

\begin{figure}
\centering
\includegraphics{figureSurf.pdf}
\caption{Response of the model to changing the ability of the generalist
\emph{and} the range of variation. While the response to environmental
change is bell-shaped when generalists are competitive (\(g\) close to
unity), it becomes monotonously decreasing for generalists with lower
competitive ability (\(g\) close to \(a\)).}\label{fig:surf}
\end{figure}

>>>>

\documentclass[10pt,a4paper,onecolumn]{article}
% \usepackage[utf8]{inputenc}
\usepackage{marginnote}
\usepackage{amssymb,amsmath}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{hyperref}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor=
{
      Timothée Poisot,
  },
            pdftitle=
{
[Re] On the coexistence of specialists and generalists
},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}
\usepackage{tcolorbox}
\usepackage{ragged2e}
\usepackage{fontspec}
\usepackage{fontawesome}
\usepackage{caption}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The coexistence of specialists and generalist within ecological
communities is a long-standing question. \textcite{wils94csg} have
suggested that this coexistence can be understood when examined in the
light of (i) differential fitness loss associated to specialism, (ii)
active habitat selection, (iii) negative density dependence due to
competition, and (iv) stochastic changes in habitat quality, that allow
combinations of species to persist even though coexistence would not be
possible in a purely deterministic world. This was an influential paper
in guiding subsequent research on the co-existence between specialists
and generalists, as it showed a strong effect on the structure of
environmental change. Here I propose an implementation of this model in
\emph{Julia} \autocite{beza17jfa}, and show that it is able to reproduce
most figures from the original manuscript.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

@@

In this matrix, \(K_1\) is the quality of habitat 1, \(K_2\) is the
quality of habitat 2, \(a\) is the fitness cost of the specialist in its
non-optimal environment, and \(b\) is the fitness cost of generalism.
Note that \(1 > b > a > 0\). The initial value of \(K_1\) is chosen
randomly, using

\[K_1 = K_{1,\text{min}} + \text{rnd}_1(K_{1,\text{max}}-K_{1,\text{min}})\,.\]

If the variation of the two environments is linked, then

\[K_2 = K_{2,\text{min}} + (1-\text{rnd}_1)(K_{2,\text{max}}-K_{2,\text{min}})\,.\]

If the variation is independent, then

\[K_2 = K_{2,\text{min}} + \text{rnd}_2(K_{2,\text{max}}-K_{2,\text{min}})\,.\]

Both \(\text{rnd}_1\) and \(\text{rnd}_2\) are random numbers uniformly
distributed in \([0,1]\).

@@

The value of \(p_l\) for which \(D_{l1}=D_{l2}\) (checked using
\textcite{maxi14mca}) is

@@

We fix \(p_m\) and \(p_n\), and find the value of \(p_l\), while
enforcing the constraint of \(0 \leq p_l \leq 1\). Iterating this
procedure a few times (10 was sufficient in all cases examined) for the
different species yields the optimal values of \(\mathbf{p}\); we can
measure the density of individuals in both habitats. Before we do so,
however, we allow a proportion \(g\) of individuals that select habitats
at random. Given a total population size of \(N_i\), there are
\(N_i(g/2)\) individuals will go to either habitat, and \(N_i(1-g)p_i\)
will pick habitat 1. With this information, we can write the matrix
describing habitat selection:

@@

The entire sequence within a timestep is: generate \(K\) for each patch;
find preferences of species for both patches; distribute the species
across patches; update population sizes based on the local fitness. This
is iterated over as many timesteps as required to reach equilibrium.

@@

\hypertarget{results}{%
\section{Results}\label{results}}

Original sources were not available, and no attempts were made to
contact the authors. For some non-stochastic situations, it is possible
to calculate expected values by hand. The original manuscript does
provide some of these values, and they were used to test this
implementation. All default parameter values are given in Table
~\ref{tbl:prm}. In all figures, circles represent the generalist, upper
and lower triangles represent the specialists of habitats 1 and 2
(respectively), and a diamond is used when both specialists have the
same behaviour.

Figure 2A and 2B in the original manuscript provide a good diagnostic
value, as they are based on non-stochastic situations. These are
reproduced in Fig. ~\ref{fig:02}. As in the original simulations, the
generalist alone overshoots its carrying capacity at \(t=5\), and the
stabilizes at a value of \(N_3^\star = 270\). Note also that
\(p_3 = 2/3\) , which represents the ratio of habitat qualities. When
adding specialists at initial densities of \(N_1 = N_2 = 1\), we observe
(i) that the preference of the generalist shifts towards habitat 1 to
avoid competition in the smallest habitat, and (ii) the density of the
specialist of habitat 2 stabilizes as soon as the generalist abandons
habitat 2 (at \(t \approx 35\)).

\hypertarget{fig:02}{%
\begin{figure}
\centering
\includegraphics{figure02.pdf}
\caption{Population dynamics and habitat preference of the generalist
alone (left), and following invasion by the two specialists at the
generalist equilibrium (right). Circles represent the generalist, upper
and lower triangles represent the specialists of habitats 1 and 2
(respectively), and a diamond is used when both specialists have the
same behaviour.}\label{fig:02}
\end{figure}
}

In Fig. ~\ref{fig:03}, this implementation gives the same qualitative
results as the original article. There are small quantitative
differences in output, which are in my opinion explained by the initial
population densities (which are not given in the original article).
Absent variation in habitat quality, generalists cannot persist, and the
specialists both reach \(N^\star = \bar K\). There is a value of
variation for which the densities of specialists and generalists are
equal (about 110 for tied variations, and 130 for independent
variations).

\hypertarget{fig:03}{%
\begin{figure}
\centering
\includegraphics{figure03.pdf}
\caption{Consequences of increasing the range of variation of both
\(K_1\) and \(K_2\). Dashed lines represent independent variation, and
solid lines tied variations, as in the original article. Circles
represent the generalist, upper and lower triangles represent the
specialists of habitats 1 and 2 (respectively), and a diamond is used
when both specialists have the same behaviour.}\label{fig:03}
\end{figure}
}

In Fig. ~\ref{fig:04}, the same qualitative result (generalists increase
in abundance when one habitat becomes larger). The quantitative
differences can, again, be explained by changes in the initial
population densities. Finally, the results in Fig. ~\ref{fig:05} and
Fig. ~\ref{fig:06} can also be replicated; it should be noted that for
Fig. ~\ref{fig:06}, the lines cross at a different point, and diverge at
different rates, than in the original paper. Additional analyses suggest
that this depends on the number of timesteps used, as well as on the
starting populations. The original manuscript does not provides these
information, and so we will have to accept a \emph{qualitative}
replication of this result.

\hypertarget{fig:04}{%
\begin{figure}
\centering
\includegraphics{figure04.pdf}
\caption{Consequences of changing the relative quality of habitats 1 and
2 on coexistence. Circles represent the generalist, upper and lower
triangles represent the specialists of habitats 1 and 2 (respectively),
and a diamond is used when both specialists have the same
behaviour.}\label{fig:04}
\end{figure}
}

\hypertarget{fig:05}{%
\begin{figure}
\centering
\includegraphics{figure05.pdf}
\caption{Consequences of changing \(b\), the ability of the generalist,
on coexistence. Dashed lines represent independent variation, and solid
lines tied variations, as in the original article. Circles represent the
generalist, upper and lower triangles represent the specialists of
habitats 1 and 2 (respectively), and a diamond is used when both
specialists have the same behaviour. Note that the direction of the axis
is \emph{reversed} with regard to the original figure.}\label{fig:05}
\end{figure}
}

\hypertarget{fig:06}{%
\begin{figure}
\centering
\includegraphics{figure06.pdf}
\caption{Consequences of changing the proportion of individuals picking
their habitat at random. Circles represent the generalist, upper and
lower triangles represent the specialists of habitats 1 and 2
(respectively), and a diamond is used when both specialists have the
same behaviour.}\label{fig:06}
\end{figure}
}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

I believe that this implementation faithfully reproduces the results of
\textcite{wils94csg}. A single run of the model (using 100 generations
as in the original article) completes in
\(\approx 5\times 10^{-3} \text{s}\). This makes this implementation
usable for teaching, as it is not time consuming to generate results. As
an example, I merged the output of Fig. ~\ref{fig:03} and Fig.
~\ref{fig:05}, to generate the response of Pielou's evenness
\autocite{piel66sps},

\[
J' = -\left(\sum_{i=1}^3\frac{N_i}{N}\text{ln}\frac{N_i}{N}\right)\frac{1}{\text{ln}\,3} \,,
\]

which measures the extent to which all species have the same density. It
appears that \(J'\), a measure of coexistence, changes in response to
both the rate of variation \emph{and} generalist ability. The results
are presented in Fig. ~\ref{fig:surf}. This figure is primarily intended
to show that this model can still generate new insights when we vary
multiple parameters: namely, the response of equitability to changes in
multiple variables is non-linear, and maximum equitability is reach for
high generalist ability but moderate range of variation (this was not
predictable from the outputs of Figs. ~\ref{fig:04} and ~\ref{fig:05}
alone).

\hypertarget{fig:surf}{%
\begin{figure}
\centering
\includegraphics{figureSurf.pdf}
\caption{Response of the model to changing the ability of the generalist
\emph{and} the range of variation. While the response to environmental
change is hump-shaped when generalists are competitive (\(b\) close to
unity), it becomes monotonously decreasing for generalists with lower
competitive ability (\(b\) close to \(a\)).}\label{fig:surf}
\end{figure}
}
